Accelerated oblique random survival forests

### 2022-06-20 paper/jmlr/main.tex
adding VI to slides

\subsection{Benchmark of prediction accuracy}

\subsubsection{Learners}

@@

Our primary metric for evaluating the accuracy of predicted risk is the integrated and scaled Brier score \citep{graf1999assessment}. Consider a testing data set:
$$\dataset_{\text{test}} = \left\{ (T_i, \delta_i, x_{i}) \right\}_{i=1}^{N_{\text{test}}}.$$
Let $\widehat{S}(t \mid x_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$.
 For observation $i$ in $\dataset_{\text{test}}$, let $\widehat{S}(t \mid \bm{x}_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$. Define \begin{align*}
\bstat = \frac{1}{\ntest} \sum_{i=1}^{\ntest} &\{ \widehat{S}(t \mid \bm{x}_i)^2 \cdot I(T_i \leq t, \delta_i = 1) \cdot \widehat{G}(T_i)^{-1} \\ &+ [1-\widehat{S}(t \mid \bm{x}_i)]^2 \cdot I(T_i > t) \cdot \widehat{G}(t)^{-1}\}
\end{align*} where $\widehat{G}(t)$ is the Kaplan-Meier estimate of the censoring distribution. As $\bstat$ is time dependent, integration over time provides a summary measure of performance over a range of plausible prediction horizons. The integrated $\bstat$ is defined as \begin{equation}
\bsbar = \frac{1}{t_2 - t_1}\int_{t_1}^{t_2} \widehat{\text{BS}}(t) dt.
\end{equation} In our results, $t_1$ and $t_2$ are the 25th and 75th percentile of event times, respectively. $\bsbar$, a sum of squared prediction errors, can be scaled to produce a measure of explained residual variation (\ie, an $R^2$ statistic) by computing \begin{equation}
R^2 = 1 - \frac{\bsbar}{\bskap}
\end{equation} where $\bskap$ is the integrated Brier score when a Kaplan-Meier estimate for survival based on the training data is used as the survival prediction function $\widehat{S}(t)$. We refer to this $R^2$ statistic as the index of prediction accuracy and we scale its values by 100 to avoid unnecessary leading zero's. For example, we present 25 if $R^2$ is 0.25 and present 10.2 if the difference between two $R^2$ is 0.102.

@@

\subsection{Benchmark of variable selection}

@@

\hspace{1em}aorsf-cph & 0.133 (0.124) & 0.757 (0.079) & 0.830 & 2.74 & 0.070 & 0.980\\
\hspace{1em}aorsf-fast & 0.133 (0.124) & 0.757 (0.080) & 0.302 & 1.00 & 0.072 & 1.00\\
\hspace{1em}aorsf-net & 0.130 (0.127) & 0.755 (0.080) & 61.237 & 202.6 & 0.069 & 0.964\\
\hspace{1em}cif-rotate & 0.125 (0.145) & 0.741 (0.092) & 30.623 & 101.3 & 7.208 & 100.3\\
\hspace{1em}cif-standard & 0.120 (0.111) & 0.748 (0.079) & 2.035 & 6.73 & 5.128 & 71.3\\
\hspace{1em}glmnet-cox & 0.119 (0.139) & 0.748 (0.083) & 0.253 & 0.837 & 0.005 & 0.070\\
\hspace{1em}rsf-standard & 0.118 (0.128) & 0.741 (0.084) & 1.694 & 5.60 & 0.152 & 2.11\\
\hspace{1em}obliqueRSF-net & 0.117 (0.094) & 0.754 (0.080) & 328.883 & 1,087.9 & 26.636 & 370.5\\
\hspace{1em}cif-extension & 0.108 (0.108) & 0.746 (0.083) & 13.510 & 44.7 & 6.840 & 95.2\\
\hspace{1em}ranger-extratrees & 0.106 (0.097) & 0.748 (0.076) & 0.692 & 2.29 & 0.264 & 3.68\\
\hspace{1em}aorsf-random & 0.101 (0.092) & 0.735 (0.074) & 1.879 & 6.22 & 0.066 & 0.919\\
\hspace{1em}xgboost-cox & 0.077 (0.114) & 0.736 (0.095) & 3.833 & 12.7 & 0.005 & 0.070\\
\hspace{1em}nn-cox & 0.030 (0.118) & 0.610 (0.128) & 15.427 & 51.0 & 1.513 & 21.0\\
\hspace{1em}xgboost-aft & -3.68 (4.86) & 0.747 (0.082) & 13.182 & 43.6 & 0.008 & 0.117\\

@@

\hspace{1em}aorsf-random & 0.028 (0.022) & 0.748 (0.038) & 0.637 & 4.19 & 0.040 & 1.06\\
\hspace{1em}ranger-extratrees & 0.028 (0.017) & 0.740 (0.036) & 0.061 & 0.404 & 0.184 & 4.81\\
\hspace{1em}obliqueRSF-net & 0.027 (0.022) & 0.746 (0.038) & 27.286 & 179.6 & 15.729 & 411.6\\
\hspace{1em}aorsf-cph & 0.025 (0.029) & 0.751 (0.042) & 0.450 & 2.97 & 0.034 & 0.891\\
\hspace{1em}cif-standard & 0.024 (0.031) & 0.744 (0.040) & 1.608 & 10.6 & 4.825 & 126.3\\
\hspace{1em}aorsf-fast & 0.024 (0.028) & 0.745 (0.044) & 0.152 & 1.00 & 0.038 & 1.00\\
\hspace{1em}cif-extension & 0.023 (0.015) & 0.722 (0.038) & 9.788 & 64.4 & 4.595 & 120.2\\
\hspace{1em}aorsf-net & 0.019 (0.034) & 0.745 (0.042) & 20.231 & 133.2 & 0.038 & 0.995\\
\hspace{1em}glmnet-cox & 0.016 (0.030) & 0.746 (0.037) & 0.183 & 1.21 & 0.002 & 0.053\\
\hspace{1em}rsf-standard & 0.005 (0.041) & 0.730 (0.042) & 0.235 & 1.55 & 0.071 & 1.85\\
\hspace{1em}cif-rotate & 0.004 (0.040) & 0.731 (0.038) & 15.089 & 99.3 & 4.246 & 111.1\\
\hspace{1em}xgboost-cox & 0.000 (0.044) & 0.751 (0.033) & 3.655 & 24.1 & 0.003 & 0.079\\
\hspace{1em}nn-cox & 0.000 (0.013) & 0.574 (0.117) & 12.345 & 81.3 & 0.639 & 16.7\\
\hspace{1em}xgboost-aft & -8.40 (1.81) & 0.737 (0.035) & 11.029 & 72.6 & 0.007 & 0.183\\

@@

\hspace{1em}obliqueRSF-net & 0.007 (0.012) & 0.819 (0.051) & 9.302 & 102.7 & 10.943 & 545.1\\
\hspace{1em}aorsf-cph & 0.007 (0.019) & 0.815 (0.058) & 0.363 & 4.00 & 0.020 & 1.00\\
\hspace{1em}aorsf-fast & 0.006 (0.019) & 0.820 (0.054) & 0.091 & 1.00 & 0.020 & 1.00\\
\hspace{1em}aorsf-random & 0.004 (0.015) & 0.780 (0.070) & 0.326 & 3.60 & 0.026 & 1.30\\
\hspace{1em}ranger-extratrees & 0.001 (0.020) & 0.773 (0.071) & 0.045 & 0.497 & 0.143 & 7.13\\
\hspace{1em}cif-extension & 0.001 (0.021) & 0.759 (0.061) & 8.354 & 92.2 & 3.905 & 194.5\\
\hspace{1em}xgboost-cox & -0.004 (0.004) & 0.500 (0.000) & 0.120 & 1.32 & 0.002 & 0.100\\
\hspace{1em}nn-cox & -0.004 (0.004) & 0.535 (0.110) & 11.322 & 125.0 & 0.689 & 34.3\\
\hspace{1em}cif-standard & -0.006 (0.026) & 0.773 (0.056) & 1.758 & 19.4 & 4.361 & 217.2\\
\hspace{1em}aorsf-net & -0.006 (0.033) & 0.799 (0.065) & 15.227 & 168.1 & 0.025 & 1.25\\
\hspace{1em}rsf-standard & -0.031 (0.053) & 0.769 (0.069) & 0.090 & 0.994 & 0.038 & 1.90\\
\hspace{1em}cif-rotate & -0.037 (0.052) & 0.704 (0.094) & 13.476 & 148.8 & 3.663 & 182.5\\
\hspace{1em}glmnet-cox & -0.062 (0.092) & 0.736 (0.098) & 0.301 & 3.33 & 0.001 & 0.051\\
\hspace{1em}xgboost-aft & -24.3 (7.36) & 0.767 (0.066) & 10.846 & 119.7 & 0.008 & 0.399\\

@@

\hspace{1em}aorsf-fast & 0.100 (0.014) & 0.718 (0.011) & 0.289 & 1.00 & 0.075 & 1.00\\
\hspace{1em}aorsf-cph & 0.100 (0.014) & 0.717 (0.011) & 0.691 & 2.39 & 0.070 & 0.945\\
\hspace{1em}cif-standard & 0.097 (0.013) & 0.710 (0.012) & 1.529 & 5.29 & 5.019 & 67.4\\
\hspace{1em}aorsf-net & 0.097 (0.014) & 0.717 (0.012) & 55.870 & 193.4 & 0.064 & 0.853\\
\hspace{1em}aorsf-random & 0.096 (0.010) & 0.716 (0.011) & 1.612 & 5.58 & 0.066 & 0.879\\
\hspace{1em}obliqueRSF-net & 0.089 (0.006) & 0.717 (0.012) & 269.561 & 932.9 & 44.243 & 593.8\\
\hspace{1em}cif-rotate & 0.087 (0.017) & 0.705 (0.014) & 15.603 & 54.0 & 6.291 & 84.4\\
\hspace{1em}rsf-standard & 0.087 (0.019) & 0.704 (0.011) & 1.646 & 5.70 & 0.143 & 1.92\\
\hspace{1em}ranger-extratrees & 0.083 (0.007) & 0.710 (0.011) & 0.735 & 2.54 & 0.263 & 3.53\\
\hspace{1em}cif-extension & 0.081 (0.006) & 0.709 (0.011) & 8.523 & 29.5 & 5.850 & 78.5\\
\hspace{1em}glmnet-cox & 0.076 (0.016) & 0.711 (0.019) & 0.132 & 0.457 & 0.004 & 0.054\\
\hspace{1em}xgboost-cox & 0.063 (0.013) & 0.701 (0.013) & 3.266 & 11.3 & 0.004 & 0.054\\
\hspace{1em}nn-cox & -0.013 (0.040) & 0.506 (0.042) & 14.970 & 51.8 & 1.780 & 23.9\\
\hspace{1em}xgboost-aft & -1.11 (0.200) & 0.706 (0.013) & 12.660 & 43.8 & 0.007 & 0.094\\

@@

\hspace{1em}aorsf-fast & 0.100 (0.017) & 0.713 (0.016) & 0.282 & 1.00 & 0.069 & 1.00\\
\hspace{1em}aorsf-cph & 0.099 (0.017) & 0.712 (0.016) & 0.708 & 2.52 & 0.070 & 1.01\\
\hspace{1em}aorsf-net & 0.096 (0.018) & 0.713 (0.017) & 51.056 & 181.3 & 0.065 & 0.937\\
\hspace{1em}cif-standard & 0.091 (0.015) & 0.701 (0.017) & 1.183 & 4.20 & 4.109 & 59.2\\
\hspace{1em}obliqueRSF-net & 0.088 (0.008) & 0.711 (0.015) & 298.417 & 1,059.9 & 44.698 & 643.9\\
\hspace{1em}aorsf-random & 0.087 (0.012) & 0.703 (0.015) & 1.629 & 5.78 & 0.064 & 0.923\\
\hspace{1em}cif-rotate & 0.084 (0.020) & 0.695 (0.017) & 14.130 & 50.2 & 6.064 & 87.4\\
\hspace{1em}cif-extension & 0.082 (0.009) & 0.707 (0.017) & 8.587 & 30.5 & 6.571 & 94.7\\
\hspace{1em}rsf-standard & 0.081 (0.020) & 0.694 (0.015) & 1.408 & 5.00 & 0.139 & 2.00\\
\hspace{1em}ranger-extratrees & 0.079 (0.011) & 0.700 (0.016) & 0.555 & 1.97 & 0.235 & 3.39\\
\hspace{1em}glmnet-cox & 0.073 (0.018) & 0.706 (0.024) & 0.161 & 0.572 & 0.005 & 0.072\\
\hspace{1em}xgboost-cox & 0.060 (0.011) & 0.695 (0.018) & 3.088 & 11.0 & 0.004 & 0.058\\
\hspace{1em}nn-cox & -0.005 (0.005) & 0.506 (0.039) & 13.814 & 49.1 & 1.499 & 21.6\\
\hspace{1em}xgboost-aft & -1.20 (0.239) & 0.701 (0.019) & 11.979 & 42.5 & 0.007 & 0.101\\

@@

\hspace{1em}obliqueRSF-net & 0.073 (0.024) & 0.749 (0.027) & 1951.148 & 2,519.7 & 15.143 & 83.8\\
\hspace{1em}cif-rotate & 0.072 (0.017) & 0.748 (0.027) & 6765.191 & 8,736.4 & 351.642 & 1,947.0\\
\hspace{1em}cif-standard & 0.068 (0.018) & 0.745 (0.029) & 8.817 & 11.4 & 4.397 & 24.3\\
\hspace{1em}cif-extension & 0.066 (0.014) & 0.746 (0.025) & 45.390 & 58.6 & 6.577 & 36.4\\
\hspace{1em}aorsf-cph & 0.065 (0.031) & 0.745 (0.024) & 1.280 & 1.65 & 0.179 & 0.992\\
\hspace{1em}aorsf-fast & 0.063 (0.031) & 0.744 (0.024) & 0.774 & 1.00 & 0.181 & 1.00\\
\hspace{1em}ranger-extratrees & 0.061 (0.024) & 0.740 (0.028) & 0.224 & 0.290 & 0.180 & 0.997\\
\hspace{1em}glmnet-cox & 0.043 (0.034) & 0.723 (0.037) & 5.608 & 7.24 & 0.005 & 0.028\\
\hspace{1em}xgboost-cox & 0.040 (0.014) & 0.743 (0.026) & 2.236 & 2.89 & 0.007 & 0.039\\
\hspace{1em}aorsf-random & 0.030 (0.015) & 0.699 (0.036) & 1.847 & 2.38 & 0.190 & 1.05\\
\hspace{1em}rsf-standard & 0.020 (0.041) & 0.693 (0.033) & 0.382 & 0.493 & 0.666 & 3.69\\
\hspace{1em}aorsf-net & 0.003 (0.067) & 0.740 (0.022) & 464.082 & 599.3 & 0.179 & 0.992\\
\hspace{1em}nn-cox & -0.008 (0.065) & 0.689 (0.045) & 18.607 & 24.0 & 1.926 & 10.7\\
\hspace{1em}xgboost-aft & -3.09 (0.664) & 0.742 (0.022) & 11.258 & 14.5 & 0.010 & 0.056\\

@@

\hspace{1em}glmnet-cox & 0.117 (0.028) & 0.787 (0.037) & 0.110 & 1.24 & 0.003 & 0.158\\
\hspace{1em}aorsf-cph & 0.101 (0.039) & 0.769 (0.034) & 0.171 & 1.93 & 0.020 & 1.05\\
\hspace{1em}aorsf-fast & 0.100 (0.038) & 0.768 (0.033) & 0.088 & 1.00 & 0.019 & 1.00\\
\hspace{1em}aorsf-net & 0.097 (0.040) & 0.760 (0.034) & 13.429 & 151.8 & 0.020 & 1.06\\
\hspace{1em}obliqueRSF-net & 0.089 (0.027) & 0.758 (0.036) & 108.537 & 1,226.6 & 5.743 & 302.0\\
\hspace{1em}cif-rotate & 0.087 (0.048) & 0.755 (0.027) & 6.229 & 70.4 & 2.161 & 113.6\\
\hspace{1em}cif-extension & 0.087 (0.036) & 0.730 (0.034) & 6.111 & 69.1 & 2.473 & 130.0\\
\hspace{1em}aorsf-random & 0.087 (0.029) & 0.757 (0.032) & 0.304 & 3.44 & 0.020 & 1.05\\
\hspace{1em}cif-standard & 0.084 (0.038) & 0.743 (0.036) & 0.793 & 8.97 & 1.275 & 67.1\\
\hspace{1em}ranger-extratrees & 0.073 (0.016) & 0.741 (0.037) & 0.036 & 0.409 & 0.087 & 4.56\\
\hspace{1em}rsf-standard & 0.072 (0.048) & 0.732 (0.034) & 0.526 & 5.94 & 0.039 & 2.07\\
\hspace{1em}xgboost-cox & 0.029 (0.050) & 0.679 (0.121) & 0.342 & 3.87 & 0.002 & 0.105\\
\hspace{1em}nn-cox & -0.006 (0.015) & 0.536 (0.097) & 12.668 & 143.2 & 0.444 & 23.3\\
\hspace{1em}xgboost-aft & -2.52 (0.477) & 0.754 (0.038) & 7.286 & 82.3 & 0.006 & 0.316\\

@@

\hspace{1em}glmnet-cox & 0.029 (0.017) & 0.620 (0.024) & 0.101 & 0.798 & 0.002 & 0.075\\
\hspace{1em}ranger-extratrees & 0.017 (0.016) & 0.596 (0.025) & 0.034 & 0.270 & 0.091 & 3.36\\
\hspace{1em}obliqueRSF-net & 0.013 (0.016) & 0.592 (0.024) & 245.138 & 1,943.8 & 11.109 & 411.1\\
\hspace{1em}aorsf-random & 0.012 (0.018) & 0.595 (0.024) & 0.473 & 3.75 & 0.024 & 0.892\\
\hspace{1em}xgboost-cox & 0.010 (0.016) & 0.598 (0.032) & 1.315 & 10.4 & 0.003 & 0.111\\
\hspace{1em}cif-standard & 0.007 (0.021) & 0.594 (0.023) & 0.950 & 7.54 & 1.722 & 63.7\\
\hspace{1em}aorsf-fast & 0.007 (0.019) & 0.594 (0.025) & 0.126 & 1.00 & 0.027 & 1.00\\
\hspace{1em}aorsf-cph & 0.007 (0.020) & 0.594 (0.026) & 0.272 & 2.16 & 0.028 & 1.04\\
\hspace{1em}aorsf-net & 0.006 (0.020) & 0.592 (0.026) & 20.063 & 159.1 & 0.027 & 1.00\\
\hspace{1em}nn-cox & 0.000 (0.019) & 0.546 (0.054) & 13.040 & 103.4 & 0.585 & 21.6\\
\hspace{1em}cif-extension & -0.005 (0.023) & 0.580 (0.028) & 6.550 & 51.9 & 3.603 & 133.3\\
\hspace{1em}cif-rotate & -0.012 (0.025) & 0.583 (0.030) & 7.129 & 56.5 & 3.081 & 114.0\\
\hspace{1em}rsf-standard & -0.027 (0.032) & 0.577 (0.024) & 1.048 & 8.31 & 0.087 & 3.22\\
\hspace{1em}xgboost-aft & -0.829 (0.337) & 0.582 (0.034) & 7.164 & 56.8 & 0.006 & 0.222\\

@@

\hspace{1em}obliqueRSF-net & 0.124 (0.016) & 0.745 (0.017) & 334.738 & 1,786.5 & 12.512 & 271.8\\

>>>>

\subsection{Benchmark of prediction accuracy and computational efficiency}

The aim of this numeric experiment is to evaluate and compare the accelerated oblique RSF with its predecessor (the oblique RSF from the \texttt{obliqueRSF} R package) and with other machine learning algorithms for risk prediction. Inferences drawn from this experiment include equivalence and inferiority tests based on Bayesian hierarchical models.

\subsubsection{Learners} \label{sec:learners}

@@

Our primary metric for evaluating the accuracy of predicted risk is the integrated and scaled Brier score \citep{graf1999assessment}. Consider a testing data set:
$$\dataset_{\text{test}} = \left\{ (T_i, \delta_i, x_{i}) \right\}_{i=1}^{N_{\text{test}}}.$$
Let $\widehat{S}(t \mid x_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$.
 For observation $i$ in $\dataset_{\text{test}}$, let $\widehat{S}(t \mid \bm{x}_i)$ be the predicted probability of survival up to a given prediction horizon of $t > 0$. Define \begin{align*}
\bstat = \frac{1}{\ntest} \sum_{i=1}^{\ntest} &\{ \widehat{S}(t \mid \bm{x}_i)^2 \cdot I(T_i \leq t, \delta_i = 1) \cdot \widehat{G}(T_i)^{-1} \\ &+ [1-\widehat{S}(t \mid \bm{x}_i)]^2 \cdot I(T_i > t) \cdot \widehat{G}(t)^{-1}\}
\end{align*} where $\widehat{G}(t)$ is the Kaplan-Meier estimate of the censoring distribution. As $\bstat$ is time dependent, integration over time provides a summary measure of performance over a range of plausible prediction horizons. The integrated $\bstat$ is defined as \begin{equation}
\bsbar = \frac{1}{t_2 - t_1}\int_{t_1}^{t_2} \widehat{\text{BS}}(t) dt.
\end{equation} In our results, $t_1$ and $t_2$ are the 25th and 75th percentile of event times, respectively. $\bsbar$, a sum of squared prediction errors, can be scaled to produce a measure of explained residual variation (\ie, an $R^2$ statistic) by computing \begin{equation}
R^2 = 1 - \frac{\bsbar}{\bskap}
\end{equation} where $\bskap$ is the integrated Brier score when a Kaplan-Meier estimate for survival based on the training data is used as the survival prediction function $\widehat{S}(t)$. We refer to this $R^2$ statistic as the index of prediction accuracy \citep{kattan2018index} and we scale its values by 100 to avoid unnecessary leading zero's. For example, we present 25 if $R^2$ is 0.25 and present 10.2 if the difference between two $R^2$ is 0.102.

Our secondary metric for evaluating predicted risk is the time-dependent concordance (C)-statistic. Briefly, we compute the time-dependent C-statistic proposed by \citet{blanche2013estimating}, which is interpreted as the probability that a risk prediction model will assign higher risk to a case (\ie, an observation with $T \leq t$ and $\delta = 1$) versus a non-case (\ie, an observation with $T > t$). Similar to the index of prediction accuracy, observations with $T \leq t$ and $\delta = 0$ only contribute to inverse proportion of censoring weights.

\subsubsection{Data sets}

We use a collection of publicly available data sets to benchmark the prediction accuracy and computational efficiency of the accelerated ORSF and each of the other learners described in \secref{sec:learners}

@@

\subsection{Benchmark of variable importance}

The aim of this numeric experiment is to compare negation VI to existing methods for VI. We consider methods that are intrinsic to the oblique RF (\eg ANOVA VI), those that are intrinsic to the RF (\eg permutation VI), and those that are model-agnostic (\eg SHAP VI). All methods are evaluated based on their ability to discriminate between variables that are related to a censored outcome versus those that are not related to it.

\subsubsection{Variable importance techniques}

We compute permutation VI using the \texttt{randomForestSRC} package. Although the \texttt{party} package implements the approach to VI developed by \citet{strobl2007bias}, the developers of the \texttt{party} package note that the implementation of this approach for survival outcomes is ``extremely slow and experimental'' as of version 1.3.10. Therefore, it is not incorporated in the current study's numeric experiments. We compute ANOVA VI using the \texttt{aorsf} package, applying a p-value threshold of 0.01 as recommended by \citet{menze2011oblique}. We compute SHAP VI for boosted tree models using the \texttt{xgboost} package and also compute SHAP VI for accelerated oblique RSFs using the \texttt{fastshap} package.

\subsubsection{Variable types}

We considered five classes of predictor variable, with each class characterized by its variables' relationship to a right-censored outcome. Specifically, \begin{itemize}
\item \textit{junk} variables had no relationship with the outcome.
\item \textit{main effect} variables were linearly related to the outcome
\item \textit{non-linear effect} variables had a sinusoidal relationship to the outcome
\item \textit{combination effect} variables had no direct relationship to the outcome. However, latent variables formed by combining the observed combination effects have a linear relationship to the outcome.
\item \textit{interaction effect} variables related to the outcome by multiplicative interaction with one other variable.
\end{itemize}

the variables have (1) no effect, (2) linear effect, (3) non-linear effect, (4) conditional linear effect, (5) combination linear effect, or (6) categorical effect on the outcome. Each conditional linear variable has an effect on the outcome that is modified by the value of one linear effect variable. Combination linear effect variables contribute to latent variables that are linearly related to the outcome. Categorical effect variables have 3 unique values, each with a different expected risk for the event of interest.

\subsubsection{Simulated data}

\subsubsection{Evaluation of variable importance}

@@

\hspace{1em}aorsf-fast & 0.132 (0.114) & 0.773 (0.071) & 0.776 & 1.00 & 0.282 & 1.00\\
\hspace{1em}rsf-standard & 0.120 (0.117) & 0.759 (0.075) & 2.016 & 2.60 & 0.444 & 1.57\\
\hspace{1em}cif-standard & 0.119 (0.101) & 0.766 (0.069) & 4.663 & 6.01 & 18.044 & 63.9\\
\hspace{1em}cif-rotate & 0.119 (0.130) & 0.757 (0.080) & 40.258 & 51.9 & 27.552 & 97.6\\
\hspace{1em}glmnet-cox & 0.114 (0.124) & 0.759 (0.076) & 0.737 & 0.949 & 0.008 & 0.028\\
\hspace{1em}xgboost-cox & 0.072 (0.103) & 0.754 (0.089) & 4.296 & 5.54 & 0.009 & 0.032\\

@@

\hspace{1em}cif-standard & 0.024 (0.031) & 0.744 (0.040) & 1.580 & 10.7 & 3.948 & 116.0\\
\hspace{1em}aorsf-fast & 0.024 (0.028) & 0.745 (0.044) & 0.147 & 1.00 & 0.034 & 1.00\\
\hspace{1em}glmnet-cox & 0.016 (0.030) & 0.746 (0.037) & 0.161 & 1.10 & 0.002 & 0.059\\
\hspace{1em}rsf-standard & 0.005 (0.041) & 0.730 (0.042) & 0.142 & 0.963 & 0.063 & 1.84\\
\hspace{1em}cif-rotate & 0.004 (0.040) & 0.731 (0.038) & 15.182 & 103.2 & 4.149 & 121.9\\
\hspace{1em}xgboost-cox & 0.000 (0.044) & 0.751 (0.033) & 3.630 & 24.7 & 0.003 & 0.088\\

@@

\hspace{1em}aorsf-fast & 0.010 (0.020) & 0.834 (0.057) & 0.088 & 1.00 & 0.020 & 1.00\\
\hspace{1em}cif-standard & -0.001 (0.025) & 0.788 (0.061) & 1.589 & 18.1 & 3.861 & 192.9\\
\hspace{1em}xgboost-cox & -0.003 (0.004) & 0.500 (0.000) & 0.112 & 1.28 & 0.002 & 0.100\\
\hspace{1em}rsf-standard & -0.022 (0.053) & 0.787 (0.071) & 0.084 & 0.960 & 0.035 & 1.75\\
\hspace{1em}cif-rotate & -0.031 (0.048) & 0.714 (0.084) & 13.331 & 152.2 & 3.293 & 164.5\\
\hspace{1em}glmnet-cox & -0.053 (0.092) & 0.736 (0.110) & 0.262 & 2.99 & 0.002 & 0.100\\
\multicolumn{7}{l}{\textit{\textbf{ARIC; coronary heart disease, n = 13626, p = 31}}}\\
\hspace{1em}aorsf-fast & 0.160 (0.008) & 0.810 (0.007) & 4.995 & 1.00 & 16.402 & 1.00\\
\hspace{1em}rsf-standard & 0.156 (0.008) & 0.803 (0.007) & 11.043 & 2.21 & 1.245 & 0.076\\
\hspace{1em}cif-standard & 0.134 (0.006) & 0.806 (0.007) & 68.008 & 13.6 & 441.507 & 26.9\\
\hspace{1em}glmnet-cox & 0.124 (0.016) & 0.793 (0.009) & 1.354 & 0.271 & 0.822 & 0.050\\
\hspace{1em}cif-rotate & 0.105 (0.003) & 0.783 (0.008) & 412.625 & 82.6 & 251.226 & 15.3\\
\hspace{1em}xgboost-cox & 0.058 (0.016) & 0.811 (0.007) & 10.324 & 2.07 & 1.073 & 0.065\\
\multicolumn{7}{l}{\textit{\textbf{ARIC; death, n = 13626, p = 31}}}\\
\hspace{1em}rsf-standard & 0.214 (0.007) & 0.785 (0.004) & 16.361 & 1.72 & 1.636 & 0.018\\
\hspace{1em}aorsf-fast & 0.211 (0.008) & 0.786 (0.005) & 9.521 & 1.00 & 92.722 & 1.00\\
\hspace{1em}cif-standard & 0.201 (0.005) & 0.784 (0.005) & 66.585 & 6.99 & 585.928 & 6.32\\
\hspace{1em}glmnet-cox & 0.181 (0.019) & 0.770 (0.008) & 2.555 & 0.268 & 1.142 & 0.012\\
\hspace{1em}cif-rotate & 0.150 (0.008) & 0.753 (0.007) & 427.333 & 44.9 & 524.889 & 5.66\\
\hspace{1em}xgboost-cox & 0.123 (0.006) & 0.789 (0.004) & 10.891 & 1.14 & 1.028 & 0.011\\
\multicolumn{7}{l}{\textit{\textbf{ARIC; heart failure, n = 13626, p = 31}}}\\
\hspace{1em}aorsf-fast & 0.236 (0.008) & 0.839 (0.006) & 5.758 & 1.00 & 21.727 & 1.00\\
\hspace{1em}rsf-standard & 0.232 (0.009) & 0.833 (0.006) & 11.647 & 2.02 & 1.259 & 0.058\\
\hspace{1em}cif-standard & 0.202 (0.006) & 0.834 (0.006) & 71.819 & 12.5 & 476.484 & 21.9\\
\hspace{1em}cif-rotate & 0.171 (0.005) & 0.805 (0.008) & 431.960 & 75.0 & 319.739 & 14.7\\
\hspace{1em}glmnet-cox & 0.163 (0.041) & 0.814 (0.016) & 2.435 & 0.423 & 1.106 & 0.051\\
\hspace{1em}xgboost-cox & 0.117 (0.019) & 0.840 (0.005) & 11.828 & 2.05 & 1.131 & 0.052\\
\multicolumn{7}{l}{\textit{\textbf{ARIC; stroke, n = 13626, p = 31}}}\\
\hspace{1em}aorsf-fast & 0.091 (0.006) & 0.791 (0.010) & 5.070 & 1.00 & 12.402 & 1.00\\
\hspace{1em}rsf-standard & 0.090 (0.007) & 0.782 (0.011) & 10.586 & 2.09 & 1.225 & 0.099\\
\hspace{1em}glmnet-cox & 0.076 (0.006) & 0.783 (0.010) & 1.135 & 0.224 & 0.905 & 0.073\\
\hspace{1em}cif-standard & 0.073 (0.005) & 0.784 (0.011) & 69.781 & 13.8 & 389.705 & 31.4\\
\hspace{1em}cif-rotate & 0.052 (0.003) & 0.766 (0.010) & 407.352 & 80.4 & 170.815 & 13.8\\
\hspace{1em}xgboost-cox & 0.048 (0.023) & 0.789 (0.010) & 7.716 & 1.52 & 0.915 & 0.074\\

@@

\hspace{1em}aorsf-fast & 0.100 (0.014) & 0.718 (0.011) & 0.259 & 1.00 & 0.075 & 1.00\\
\hspace{1em}cif-standard & 0.097 (0.013) & 0.710 (0.012) & 1.193 & 4.60 & 3.848 & 51.3\\
\hspace{1em}cif-rotate & 0.087 (0.017) & 0.705 (0.014) & 13.358 & 51.5 & 5.116 & 68.1\\
\hspace{1em}rsf-standard & 0.087 (0.019) & 0.704 (0.011) & 1.226 & 4.73 & 0.150 & 2.00\\
\hspace{1em}glmnet-cox & 0.076 (0.016) & 0.711 (0.019) & 0.116 & 0.448 & 0.004 & 0.053\\
\hspace{1em}xgboost-cox & 0.063 (0.013) & 0.701 (0.013) & 3.165 & 12.2 & 0.004 & 0.053\\

@@

\hspace{1em}aorsf-fast & 0.100 (0.017) & 0.713 (0.016) & 0.259 & 1.00 & 0.076 & 1.00\\
\hspace{1em}cif-standard & 0.091 (0.015) & 0.701 (0.017) & 1.182 & 4.56 & 3.929 & 51.7\\
\hspace{1em}cif-rotate & 0.084 (0.020) & 0.695 (0.017) & 13.338 & 51.5 & 5.118 & 67.3\\
\hspace{1em}rsf-standard & 0.081 (0.020) & 0.694 (0.015) & 1.223 & 4.72 & 0.151 & 1.99\\
\hspace{1em}glmnet-cox & 0.073 (0.018) & 0.706 (0.024) & 0.115 & 0.444 & 0.004 & 0.053\\
\hspace{1em}xgboost-cox & 0.060 (0.011) & 0.695 (0.018) & 2.895 & 11.2 & 0.004 & 0.053\\

@@

\hspace{1em}cif-rotate & 0.070 (0.018) & 0.747 (0.027) & 6304.881 & 8,318.3 & 374.594 & 2,134.8\\
\hspace{1em}cif-standard & 0.067 (0.019) & 0.747 (0.030) & 8.059 & 10.6 & 4.493 & 25.6\\
\hspace{1em}aorsf-fast & 0.065 (0.028) & 0.746 (0.026) & 0.758 & 1.00 & 0.175 & 1.00\\
\hspace{1em}glmnet-cox & 0.041 (0.032) & 0.724 (0.036) & 5.686 & 7.50 & 0.005 & 0.029\\
\hspace{1em}xgboost-cox & 0.027 (0.034) & 0.741 (0.030) & 2.154 & 2.84 & 0.006 & 0.034\\
\hspace{1em}rsf-standard & 0.024 (0.037) & 0.695 (0.033) & 0.323 & 0.427 & 0.666 & 3.80\\

@@

\hspace{1em}glmnet-cox & 0.117 (0.028) & 0.787 (0.037) & 0.082 & 0.988 & 0.002 & 0.111\\
\hspace{1em}aorsf-fast & 0.100 (0.038) & 0.768 (0.033) & 0.083 & 1.00 & 0.018 & 1.00\\
\hspace{1em}cif-rotate & 0.087 (0.048) & 0.755 (0.027) & 6.369 & 76.7 & 1.766 & 98.0\\
\hspace{1em}cif-standard & 0.084 (0.038) & 0.743 (0.036) & 0.751 & 9.04 & 1.020 & 56.6\\
\hspace{1em}rsf-standard & 0.072 (0.048) & 0.732 (0.034) & 0.589 & 7.08 & 0.040 & 2.22\\
\hspace{1em}xgboost-cox & 0.029 (0.050) & 0.679 (0.121) & 0.320 & 3.86 & 0.002 & 0.111\\

@@

\hspace{1em}glmnet-cox & 0.029 (0.017) & 0.620 (0.024) & 0.086 & 0.717 & 0.003 & 0.120\\
\hspace{1em}xgboost-cox & 0.010 (0.016) & 0.598 (0.032) & 1.272 & 10.6 & 0.003 & 0.120\\
\hspace{1em}cif-standard & 0.007 (0.021) & 0.594 (0.023) & 0.753 & 6.27 & 1.153 & 46.1\\
\hspace{1em}aorsf-fast & 0.007 (0.019) & 0.594 (0.025) & 0.120 & 1.00 & 0.025 & 1.00\\
\hspace{1em}cif-rotate & -0.012 (0.025) & 0.583 (0.030) & 7.360 & 61.3 & 2.955 & 118.1\\
\hspace{1em}rsf-standard & -0.027 (0.032) & 0.577 (0.024) & 0.742 & 6.18 & 0.092 & 3.68\\

@@

\hspace{1em}cif-standard & 0.123 (0.020) & 0.742 (0.019) & 0.919 & 5.21 & 2.297 & 52.1\\
